NumPy---Array Computation for Python

### 2019-12-16 paper.tex
NEW: Draft of section for `__array_function__`.

For many years, NumPy operated without external funding, development
time mostly being contributed freely by students and researchers in
their spare time.  In-person meetings were sponsored off of
individual's research grants, and while industry sponsored
contributions were made, notably those by Mark Wiebe when he was
sponsored by Travis Oliphant at Continuum Analytics, mostly the package had rapid growth
and adoption without external investment.  In 2014, NumPy received its
first large grants from the Moore and Sloan foundations, totaling 1.3M
dollars.  This allowed for hiring several full time programmers and,
with the support of community members, mobilized a project that had
become stagnant for the several years prior.

NumPy is now actively developed, with weekly online developer
meetings, a formalized NumPy enhancement proposal process, and
multiple active grants (but still almost no industry support).

@@

\subsection*{Project goals / scope of library}

\subsection*{Current status (maturity, users)}

\subsection*{Enabled ecosystem }

% NEP 29 (which releases are supported in rolling window)

% NumPyDoc standard, recent improvements

@@

% Wheels build system

@@

\subsection*{NEPS revived and roadmap}

@@

\subsection*{\_\_array\_func\_\_}

@@

@@

% BIDS -- UCB, CZI -- NumFOCUS/QuanSight, Tidelift, ...

@@

\subsection*{Maintainer overview}

\subsection*{Industry collaboration}

@@

% The Discussion should be succinct and must not contain subheadings.

% \subsubsection*{Impact Now}

@@

% \subsubsection*{Future Development}

\section*{Data Availability}
% The statement should be provided as a separate section (titled 'Data Availability') at the end of the main text, before the 'References' section.

@@

%To include, in this order: \textbf{Accession codes} (where applicable);
%\textbf{Competing financial interests} (mandatory statement).

%The corresponding author is responsible for submitting a
%\href{http://www.nature.com/srep/policies/index.html#competing}{competing
%financial interests statement} on behalf of all authors of the paper. This
%statement must be included in the submitted article file.

>>>>

@@

For many years, NumPy operated without external funding, development
time mostly being contributed freely by students and researchers in
their spare time.  In-person meetings were sponsored off of
individual's research grants, and while industry sponsored
contributions were made, notably those by Mark Wiebe when he was
sponsored by Travis Oliphant at Continuum Analytics, mostly the package had rapid growth
and adoption without external investment.

% In 2014, NumPy received its
% first large grants from the Moore and Sloan foundations, totaling 1.3M
% dollars.  This allowed for hiring several full time programmers and,
% with the support of community members, mobilized a project that had
% become stagnant for the several years prior.

% BIDS -- UCB
% https://www.moore.org/grant-detail?grantId=GBMF5447
% $645,020 in 2016
% https://sloan.org/grant-detail/8222
% $659,359 in 2017
% https://bids.berkeley.edu/news/bids-receives-sloan-foundation-grant-contribute-numpy-development

The Berkeley Institute for Data Science (BIDS) received two grants to support
the development of NumPy.
At the end of 2016, the Gordon and Betty Moore Foundation awarded them
\$645,020 for improving NumPy for better data science.
In 2017, the Alfred P. Sloan Foundation complemented the funding with an
additional \$659,359.
Stéfan van der Walt, a senior research data scientist at BIDS and a member of
the NumPy steering committee, is the PI of both grants and manages a team of
four programmers working on the project.

% how much do we say about who these developers are and
% lessons learned?

% technical focus 
% improve various aspects of NumPy, including easier construction of custom data types (e.g., missing values, physical units, times & dates) and better integration with arrays customized for specialized domains (e.g., SciPy's sparse arrays, pandas' DataFrames, and xarray's labeled arrays). 
% technical work, and it increased the velocity of the projects by ~25-30%, and in addition it enabled integrating some larger changes, like the numpy.random redesign.

% CZI -- NumFOCUS/QuanSight
% https://chanzuckerberg.com/eoss/proposals/strengthening-numpys-foundations-growing-beyond-code/
% NumPy and OpenBLAS received $195,000 in 2019
% https://labs.quansight.org/blog/2019/11/numpy-openblas-CZI-grant/

NumFOCUS received a grant for \$195,000 in 2019 for the development of NumPy
and OpenBLAS from the Chan Zuckerberg Initiative. 
NumPy depends on OpenBLAS to provide its underlying accelerated linear algebra implementation.
Ralf Gommers, the Director of Quansight Labs and a member of
the NumPy steering committee, is the PI.

% focus on better serve NumPy's large number of beginner to intermediate level users
% focusing all our time on growing the team and better serving new users and contributors

% Tidelift, ...
% https://tidelift.com/subscription/pkg/pypi-numpy
% $10,000 over 24 months according to 
% https://mail.python.org/pipermail/numpy-discussion/2018-September/078736.html

\subsection*{Current status}

NumPy is now actively developed, with weekly online developer
meetings, a formalized NumPy enhancement proposal process, and
multiple active grants (but still almost no industry support).
% check almost no industry

% (maturity, users)

\subsubsection*{Project goals / scope of library}

% https://numpy.org/neps/scope.html

\subsubsection*{NEPS revived and roadmap}

% https://numpy.org/neps/roadmap.html

@@

\subsubsection*{Wheels build system}

A Python \emph{wheel}\cite{PEP427} is a standard file format for
distributing Python libraries.  In addition to Python code, a
wheel may include compiled C extensions and other binary data.
This is important, because many libraries, including NumPy,
require a C compiler and other build tools to build the software
from the source code, making it difficult for many users to install
the software on their own.  The introduction of wheels to the Python
packaging system has made it much easier for users to install
precompiled libraries.

A GitHub repository containing scripts to build NumPy wheels has
been configured so that a simple commit to the repository triggers
an automated build system that creates NumPy wheels for several
computer platforms, including Windows, Mac OSX and Linux.  The wheels
are uploaded to a public server and made available for anyone to use.
This system makes it easy for users to install precompiled versions
of NumPy on these platforms.

The technology that is used to build the wheels evolves continually.
At the time this paper is being written, a key component is the
\code{multibuild} suite of tools developed by Matthew Brett and
other developers\cite{multibuild}.  Currently, scripts using
\code{multibuild} are written for the continuous integration
platforms Travis-CI (for Linux and Mac OSX) and Appveyor
(for Windows).

\subsubsection*{NEP 29}

% https://numpy.org/neps/nep-0029-deprecation_policy.html
In NEP 29, we recommend, along with leaders from various other
projects, that all projects across the Scientific Python ecosystem
adopt a common ``time window-based'' policy for support of Python and
NumPy versions. This standard will simplify downstream project and
release planning.

\subsubsection*{NumPyDoc standard}

% recent improvements

@@

@@

\subsection*{The \_\_array\_function\_\_ Protocol}

The central position which NumPy takes within the python ecosystem with
respect to scientific computing and numerical works,
means that a vast number of projects are build on it.
These projects are consumers of the NumPy API.

On the other side, projects are providers of a
\emph{NumPy-like API} and array objects targeting audiences with specialized needs beyond
NumPy's capabilities. Some of these projects provide:

% TODO: A list prints too large currently, so probably should make this
%       text (although depends a bit on the final layout maybe)
\begin{description}
    \item[SciPy's \code{scipy.sparse} and \code{pydata.sparse}:] Sparse arrays
    \item[CuPy:] Computation on the GPU
    \item[Dask:] Parallel, distributed computing
    \item[PyTorch and TensorFlow:] Deep learning frameworks.
\end{description}

The multitude of specialized projects creates the difficulty that consumers
of these NumPy-like APIs write code specific to a single project and do not support
all of the above array providers.
This is a burden for users relying on the specialized Array-like, since
a tool they need may not work for them.
Additionally, a user who has the need of transitioning from NumPy to a more
specialized array, is faced with a challenge.
These issue can lead into a fragmentation of the user bases into consumers
of different NumPy-like APIs.

To address these issues NumPy has the goal of providing the fundamental
API for \emph{interoperability} between the various NumPy-like APIs.
An earlier step in this direction was the implementation of the
\code{\_\_array\_ufunc\_\_} protocol in NumPy 1.13, which enabled interoperability
for most mathematical functions.\cite{NEP13}
In 2019 this was expanded more generally with the inclusion of the
\code{\_\_array\_function\_\_} protocol into NumPy~1.17.
These two protocols allows providers of Array-like APIs to be interoperable
with the NumPy API: their arrays work correctly with almost all NumPy functions.\cite{NEP18}
For the users relying on specialized array projects it means that even though
much code is written specifically for NumPy arrays and uses the NumPy API as
\code{import numpy as np}, it can nevertheless work for them.
A brief example of such a function which is written using NumPy
but can now be used more generally is:
\begin{lstlisting}
def f(x):
    y = np.tensordot(x, x.T)
    return np.mean(np.exp(y))
\end{lstlisting}
For example, a user relying on this function and normally using it with a
NumPy array, may now use a distributed Dask array when challenged
by a very large dataset, without any modification to \code{f(x)}.

While, as of writing this, the interface is still considered experimental
and may be modified, it has been implemented by multiple projects already
and proven useful.
% TODO: Should likely have citation/example if I claim it was useful (based
%       mainly on Matt's statement).

@@

% incl. CoC

@@

% may be useful:
% - https://github.com/docathon/watchtower
% - https://github.com/machine-shop/oss_devkit/blob/master/plot_pr.py % (PR over time plot, there may be other better versions out there)

@@

@@

\subsubsection*{Impact Now}

@@

\subsubsection*{Future Development}

% Dtypes

The meaning of each element within a NumPy array is described by its datatype
(dtype). NumPy arrays can hold all common numerical datatypes, string, datetimes
and generic Python objects, each of these is identified by a dtype.
We are busy overhauling the datatype system to make it consistent (aligned with
expectations), to simplify the creation of custom dtypes.
While numerical dtypes make the foundation for most users, there are
a multitude of use cases which cannot prosper due to current limitations.
These use cases include datatypes with physical units\cite{astropy,Goldbaum2018,pint},
% pyadolc may be a bit too small a project, so may want to remove/replace with an other example.
geometrical objects\cite{pygeos}, automatic differentiation\cite{pyadolc}, and many more.
Projects addressing these exist, each struggles with current
limitations.  % TODO: I may need citations, we could link github issues.

While it is currently possible to define custom dtypes within NumPy, many of the
use cases above cannot be solved using this system.
This is evidenced also by code within NumPy which would require changes
in many places to add such dtypes, even when adding them within NumPy itself.
The new implementation will be simpler and more cleanly implemented in
this regard.
Custom dtypes and the basic dtypes included within NumPy will be implemented
and behave identically as much as possible.
At the same time redesigning the API will make NumPy dtypes more flexible to grow
to future needs should they arise.

As a further step, to make dtypes more accessible and allow a faster adoption of
custom dtypes within the community, we plan to create a Python API to
define such custom dtypes.

A \href{https://github.com/numpy/numpy/pull/14422}{draft of DTypes NEP}
is under development in consultation with the community.

\section*{Data Availability}

@@
